Data Analysis

### 2023-03-30 ms1_data_analysis.Rmd
Before removing the focused model

library(ggplot2)
library(dplyr)

@@

##### Change in Biomass: log-respose ratio of a species' biomass

@@

# Create data-set for a given treatment
deltaData <- data.frame()

# Change total bio variable into numeric
lgsData$tot_bio <- as.numeric(lgsData$tot_bio)

# Loop through each species in each garden and site and calculate their deltaBio
for (st in unique(lgsData$site)){

  for (gard in unique(lgsData$garden)){

    stGardDat <- lgsData %>%
      filter(site == st,
             garden == gard)

    if (dim(stGardDat)[1] == 0){
      # If subset is empty skip to another one
      next
    }

    if (length(unique(stGardDat$treatment)) < 2){
      # If there is species in only one treatment skip to another one
      next
    }

    specInCtr <- unique(stGardDat[stGardDat$treatment == subsetVars[1], ]$unified_names)
    specInTrt <- unique(stGardDat[stGardDat$treatment == subsetVars[2], ]$unified_names)
    specInBoth <- specInCtr[specInCtr %in% specInTrt]

    cDat <- stGardDat %>%
      filter(unified_names %in% specInBoth,
             treatment == subsetVars[1]) %>%
      arrange(unified_names)
    tDat <- stGardDat %>%
      filter(unified_names %in% specInBoth,
             treatment == subsetVars[2]) %>%
      arrange(unified_names)

    # Calculate LRR for biomass and SLA change

    cDat$comparison <- subsetVars[2]

    if (subsetVars[2] == "h"){
      cDat$lrrtb <- log(tDat$tot_bio/cDat$tot_bio)
      cDat$lrrsla <- log(tDat$sla.d.m2kg1/cDat$sla.d.m2kg1)
    } else{
      cDat$lrrtb <- log(cDat$tot_bio/tDat$tot_bio)
      cDat$lrrsla <- log(cDat$sla.d.m2kg1/tDat$sla.d.m2kg1)
    }

    deltaData <- rbind(deltaData, cDat)

# Log transform estimated leaf area
deltaData <- deltaData %>%
          mutate(lest.leaf.area = log(est.leaf.area))%>%
  filter(!is.infinite(lest.leaf.area))

@@

# Log-response ratios of total biomass vs of sla

# DeltaBiomass vs SLA at each site
slaPlot <- ggplot(deltaData, aes(x = lsla.d, y = lrrtb, color = site)) +
  geom_point()+
  geom_smooth(method = "lm", formula = y ~ x + I(x^2))

# There might be some effect of site.
# DeltaBiomass vs SLA at each site
wcPlot <- ggplot(deltaData, aes(x = water.cont, y = lrrtb, color = site)) +
  geom_point()+
  geom_smooth(method = "lm", formula = y ~ x + I(x^2))
# Seems like the effect is not very strong either

herbPlot <- ggplot(deltaData, aes(x = perc.herb, y = lrrtb, color = site)) +
  geom_point()+
  geom_smooth(method = "lm", formula = y ~ x + I(x^2)) + 
  scale_x_continuous(trans = 'log2')

```
It seems that in some cases the nonlinear relationship may be important. Generally water content looks linear for each elevation. We will include this and build also a model that models log-ratio biomass vs water content as a linear relationship. We arrive to the final definition of the full model.

@@

```{r}
# Model explorations
simple.sla <- gls(
  lrrtb ~ lsla.d*site + I(lsla.d^2)*site + richness + labu + lrrsla,
                data = deltaData %>% filter(unified_names %in% myTree.rand.poly$tip.label),
  method = "ML"
  )

@@

Table. Results of the selected model for the deltaBio.

```{r}
sjPlot::tab_model(full.sla.linear.noint.nosite)
```

@@

# Wanang ----
wdata <- deltaData %>% 
  filter(unified_names %in% myTree.rand.poly$tip.label) %>%
  filter(site == "wanang")

# dd.gls.focsed.w <- nlme::gls(lrrtb ~ lsla.d,
#                      data = wdata,
#                      method = "ML",
#                      correlation = corMartins(2.718282, 
#                                               phy = myTree.rand.poly, 
#                                               form = ~unified_names,
#                                               fixed = T),
#                      na.action = na.omit)
dd.gls.focsed.w <- nlme::gls(lrrtb ~ lsla.d + water.cont + perc.herb + richness + labu + lrrsla,
                     data = wdata,
                     method = "ML",
                     correlation = corMartins(2.718282, 
                                              phy = myTree.rand.poly, 
                                              form = ~unified_names,
                                              fixed = T),
                     na.action = na.omit)

@@

# Predict values
pred.vals <- AICcmodavg::predictSE.gls(dd.gls.focsed.w, wdata, 
                                       se.fit = T)
wdata <- wdata %>%
  mutate(preds = pred.vals$fit,
         se.preds = pred.vals$se.fit,
         se.upr = preds + se.preds,
         se.lwr = preds - se.preds)

# print the MSE
mse_focused_df <- data.frame(mse = with(wdata, mean((lrrtb - preds))^2),
                             treatment = "Insecticide", site = "Wanang")

# Numba ----
ndata <- deltaData %>% 
  filter(unified_names %in% myTree.rand.poly$tip.label) %>%
  filter(site == "numba")

# dd.gls.focsed.n <- nlme::gls(lrrtb ~ lsla.d,
#                      data = ndata,
#                      method = "ML",
#                      correlation = corMartins(2.718282, 
#                                               phy = myTree.rand.poly, 
#                                               form = ~unified_names,
#                                               fixed = T),
#                      na.action = na.omit)
dd.gls.focsed.n <- nlme::gls(lrrtb ~ lsla.d + water.cont + perc.herb + richness + labu + lrrsla,
                     data = ndata,
                     method = "ML",
                     correlation = corMartins(2.718282, 
                                              phy = myTree.rand.poly, 
                                              form = ~unified_names,
                                              fixed = T),
                     na.action = na.omit)

# Predict values
pred.vals <- AICcmodavg::predictSE.gls(dd.gls.focsed.n, ndata, 
                                       se.fit = T)
ndata <- ndata %>%
  mutate(preds = pred.vals$fit,
         se.preds = pred.vals$se.fit,
         se.upr = preds + se.preds,
         se.lwr = preds - se.preds)

# print the MSE
mse_focused_df <- rbind(mse_focused_df,  
                        data.frame(mse = with(ndata, mean((lrrtb - preds))^2),
                                   treatment = "Insecticide", site = "Numba"))

# Yawan ----
ydata <- deltaData %>% 
  filter(unified_names %in% myTree.rand.poly$tip.label) %>%
  filter(site == "yawan")

@@

dd.gls.focsed.y <- nlme::gls(lrrtb ~ lsla.d + water.cont + perc.herb + richness + labu + lrrsla,
                     data = ydata,
                     method = "ML",
                     correlation = corMartins(2.718282,
                                              phy = myTree.rand.poly,
                                              form = ~unified_names,
                                              fixed = T),
                     na.action = na.omit)

# Predict values
pred.vals <- AICcmodavg::predictSE.gls(dd.gls.focsed.y, ydata, 
                                       se.fit = T)
ydata <- ydata %>%
  mutate(preds = pred.vals$fit,
         se.preds = pred.vals$se.fit,
         se.upr = preds + se.preds,
         se.lwr = preds - se.preds)

# print the MSE
mse_focused_df <- rbind(mse_focused_df,  
                        data.frame(mse = with(ydata, mean((lrrtb - preds))^2),
                                   treatment = "Insecticide", site = "Yawan"))

library(sjPlot)

tab_model(dd.gls.focsed.w, dd.gls.focsed.n, dd.gls.focsed.y)

# fig.cap = "Empirical and predicted values of a simple model of biomass change vs logarithm of SLA for each studied sites."}

# Plot the relationship and add predicted values
wpredPlot <- ggplot(wdata, aes(y = lrrtb, x = lsla.d))+
  geom_point() +
  geom_line(data = wdata, aes(x = lsla.d, y = preds), lwd = 1,
            lty = 2)+
  # geom_ribbon(aes(ymin=se.lwr, ymax=se.upr), alpha = .15) +
  ggtitle("Wanang 200 m a.s.l.")

npredPlot <- ggplot(ndata, aes(y = lrrtb, x = lsla.d))+
  geom_point() +
  geom_line(data = ndata, aes(x = lsla.d, y = preds), 
            lwd = 1, lty = 2)+
  # geom_ribbon(aes(ymin=se.lwr, ymax=se.upr), alpha = .15) + 
  ggtitle("Numba 750 m a.s.l.")

ypredPlot <- ggplot(ydata, aes(y = lrrtb, x = lsla.d))+
  geom_point() +
  geom_line(data = ydata, aes(x = lsla.d, y = preds), lwd = 1)+
  geom_ribbon(aes(ymin=se.lwr, ymax=se.upr), alpha = .15) + 
  ggtitle("Yawan 1900 m a.s.l.")

ggpubr::ggarrange(wpredPlot, npredPlot, ypredPlot,
                  nrow = 1)
# Plot the relationship and add predicted values
ggplot(rbind(ydata, ndata, wdata), aes(y = lrrtb, x = lsla.d, color = site))+
  geom_point() + 
  geom_line(data = rbind(ydata, ndata, wdata), 
            aes(x = lsla.d, y = preds), lwd = 1)+
  geom_ribbon(aes(ymin=se.lwr, ymax=se.upr), alpha = .15)

sjPlot::tab_df(mse_focused_df)

@@

# Fungicide

@@

# Additional herbivores

@@

>>>>

library(ggplot2)
library(dplyr)
library(sjPlot)

@@

@@

# # Create data-set for a given treatment
# deltaData <- data.frame()
# 
# # Change total bio variable into numeric
# lgsData$tot_bio <- as.numeric(lgsData$tot_bio)
# 
# # Loop through each species in each garden and site and calculate their deltaBio
# for (st in unique(lgsData$site)){
#   
#   for (gard in unique(lgsData$garden)){
#     
#     stGardDat <- lgsData %>%
#       filter(site == st,
#              garden == gard)
#     
#     
#     if (dim(stGardDat)[1] == 0){
#       # If subset is empty skip to another one
#       next
#     }
#     
#     if (length(unique(stGardDat$treatment)) < 2){
#       # If there is species in only one treatment skip to another one
#       next
#     }
#     
#     specInCtr <- unique(stGardDat[stGardDat$treatment == subsetVars[1], ]$unified_names)
#     specInTrt <- unique(stGardDat[stGardDat$treatment == subsetVars[2], ]$unified_names)
#     specInBoth <- specInCtr[specInCtr %in% specInTrt]
#     
#     cDat <- stGardDat %>%
#       filter(unified_names %in% specInBoth,
#              treatment == subsetVars[1]) %>%
#       arrange(unified_names)
#     tDat <- stGardDat %>%
#       filter(unified_names %in% specInBoth,
#              treatment == subsetVars[2]) %>%
#       arrange(unified_names)
#     
#     # Calculate LRR for biomass and SLA change
#     
#     cDat$comparison <- subsetVars[2]
#     
#     if (subsetVars[2] == "h"){
#       cDat$lrrtb <- log(tDat$tot_bio/cDat$tot_bio)
#       cDat$lrrsla <- log(tDat$sla.d.m2kg1/cDat$sla.d.m2kg1)
#     } else{
#       cDat$lrrtb <- log(cDat$tot_bio/tDat$tot_bio)
#       cDat$lrrsla <- log(cDat$sla.d.m2kg1/tDat$sla.d.m2kg1)
#     }
#     
#     deltaData <- rbind(deltaData, cDat)
#     
#   }
# }
# 
# 
# # Log transform estimated leaf area
# deltaData <- deltaData %>%
#           mutate(lest.leaf.area = log(est.leaf.area))%>%
#   filter(!is.infinite(lest.leaf.area))

@@

It seems that in some cases the nonlinear relationship may be important. Generally water content looks linear for each elevation. We will include this and build also a model that models log-ratio biomass vs water content as a linear relationship. We arrive to the final definition of the full model.

@@

# Model explorations
simple.sla <- gls(
  lrrtb ~ lsla.d*site + I(lsla.d^2)*site + richness + labu + lrrsla,
                data = deltaData %>% filter(unified_names %in% myTree.rand.poly$tip.label),
  method = "ML"
  )

@@

Table. Results of the selected model for the deltaBio.
```{r}
sjPlot::tab_model(full.sla.linear.noint.nosite)
```

@@

# Wanang ----
# wdata <- deltaData %>% 
#   filter(unified_names %in% myTree.rand.poly$tip.label) %>%
#   filter(site == "wanang")

# dd.gls.focsed.w <- nlme::gls(lrrtb ~ lsla.d,
#                      data = wdata,
#                      method = "ML",
#                      correlation = corMartins(2.718282, 
#                                               phy = myTree.rand.poly, 
#                                               form = ~unified_names,
#                                               fixed = T),
#                      na.action = na.omit)
# dd.gls.focsed.w <- nlme::gls(lrrtb ~ lsla.d + water.cont + perc.herb + richness + labu + lrrsla,
#                      data = wdata,
#                      method = "ML",
#                      correlation = corMartins(2.718282, 
#                                               phy = myTree.rand.poly, 
#                                               form = ~unified_names,
#                                               fixed = T),
#                      na.action = na.omit)

@@

# Predict values
# pred.vals <- AICcmodavg::predictSE.gls(dd.gls.focsed.w, wdata, 
#                                        se.fit = T)
# wdata <- wdata %>%
#   mutate(preds = pred.vals$fit,
#          se.preds = pred.vals$se.fit,
#          se.upr = preds + se.preds,
#          se.lwr = preds - se.preds)
# 
# # print the MSE
# mse_focused_df <- data.frame(mse = with(wdata, mean((lrrtb - preds))^2),
#                              treatment = "Insecticide", site = "Wanang")

# Numba ----
# ndata <- deltaData %>% 
#   filter(unified_names %in% myTree.rand.poly$tip.label) %>%
#   filter(site == "numba")

# dd.gls.focsed.n <- nlme::gls(lrrtb ~ lsla.d,
#                      data = ndata,
#                      method = "ML",
#                      correlation = corMartins(2.718282, 
#                                               phy = myTree.rand.poly, 
#                                               form = ~unified_names,
#                                               fixed = T),
#                      na.action = na.omit)
# dd.gls.focsed.n <- nlme::gls(lrrtb ~ lsla.d + water.cont + perc.herb + richness + labu + lrrsla,
#                      data = ndata,
#                      method = "ML",
#                      correlation = corMartins(2.718282, 
#                                               phy = myTree.rand.poly, 
#                                               form = ~unified_names,
#                                               fixed = T),
#                      na.action = na.omit)

# Predict values
# pred.vals <- AICcmodavg::predictSE.gls(dd.gls.focsed.n, ndata, 
#                                        se.fit = T)
# ndata <- ndata %>%
#   mutate(preds = pred.vals$fit,
#          se.preds = pred.vals$se.fit,
#          se.upr = preds + se.preds,
#          se.lwr = preds - se.preds)
# 
# # print the MSE
# mse_focused_df <- rbind(mse_focused_df,  
#                         data.frame(mse = with(ndata, mean((lrrtb - preds))^2),
#                                    treatment = "Insecticide", site = "Numba"))

# Yawan ----
# ydata <- deltaData %>% 
#   filter(unified_names %in% myTree.rand.poly$tip.label) %>%
#   filter(site == "yawan")

@@

# dd.gls.focsed.y <- nlme::gls(lrrtb ~ lsla.d + water.cont + perc.herb + richness + labu + lrrsla,
#                      data = ydata,
#                      method = "ML",
#                      correlation = corMartins(2.718282,
#                                               phy = myTree.rand.poly,
#                                               form = ~unified_names,
#                                               fixed = T),
#                      na.action = na.omit)

# Predict values
# pred.vals <- AICcmodavg::predictSE.gls(dd.gls.focsed.y, ydata, 
#                                        se.fit = T)
# ydata <- ydata %>%
#   mutate(preds = pred.vals$fit,
#          se.preds = pred.vals$se.fit,
#          se.upr = preds + se.preds,
#          se.lwr = preds - se.preds)
# 
# # print the MSE
# mse_focused_df <- rbind(mse_focused_df,  
#                         data.frame(mse = with(ydata, mean((lrrtb - preds))^2),
#                                    treatment = "Insecticide", site = "Yawan"))

# tab_model(dd.gls.focsed.w, dd.gls.focsed.n, dd.gls.focsed.y)
# 
# 
# # fig.cap = "Empirical and predicted values of a simple model of biomass change vs logarithm of SLA for each studied sites."}
# 
# # Plot the relationship and add predicted values
# wpredPlot <- ggplot(wdata, aes(y = lrrtb, x = lsla.d))+
#   geom_point() +
#   geom_line(data = wdata, aes(x = lsla.d, y = preds), lwd = 1,
#             lty = 2)+
#   # geom_ribbon(aes(ymin=se.lwr, ymax=se.upr), alpha = .15) +
#   ggtitle("Wanang 200 m a.s.l.")
# 
# npredPlot <- ggplot(ndata, aes(y = lrrtb, x = lsla.d))+
#   geom_point() +
#   geom_line(data = ndata, aes(x = lsla.d, y = preds), 
#             lwd = 1, lty = 2)+
#   # geom_ribbon(aes(ymin=se.lwr, ymax=se.upr), alpha = .15) + 
#   ggtitle("Numba 750 m a.s.l.")
# 
# ypredPlot <- ggplot(ydata, aes(y = lrrtb, x = lsla.d))+
#   geom_point() +
#   geom_line(data = ydata, aes(x = lsla.d, y = preds), lwd = 1)+
#   geom_ribbon(aes(ymin=se.lwr, ymax=se.upr), alpha = .15) + 
#   ggtitle("Yawan 1900 m a.s.l.")
# 
# ggpubr::ggarrange(wpredPlot, npredPlot, ypredPlot,
#                   nrow = 1)
# # Plot the relationship and add predicted values
# ggplot(rbind(ydata, ndata, wdata), aes(y = lrrtb, x = lsla.d, color = site))+
#   geom_point() + 
#   geom_line(data = rbind(ydata, ndata, wdata), 
#             aes(x = lsla.d, y = preds), lwd = 1)+
#   geom_ribbon(aes(ymin=se.lwr, ymax=se.upr), alpha = .15)
# 
# sjPlot::tab_df(mse_focused_df)

@@

#### Fungicide

@@

#### Additional herbivores

@@

#### Predators

```{r}
subsetVars <- c("c", "p")

source('code/model_definitions_pgls.R')

# Model selection.

# 1. Full model (quadratic, and site interactions) vs no traits
anova(no.traits, full.model)            # full model is better

# 2. Removal of individual traits

# 2.1 Remove SLA
anova(full.model, q.herb.q.water) # withouth sla is slightly better (AIC)

# 2.2 Remove herbivory
anova(full.model, q.sla.q.water.noint) # Model without herbivory is not better

# 2.3 Remove water content?
anova(full.model, q.sla.q.herb) # Model with herbivory is much better

# 3. Reduce model even further - 1 trait models.
anova(full.model, just.sla.quadratic)
anova(full.model, just.herb.quadratic)
anova(full.model, just.water.quadratic)

# Seems like they all stay!

# 4. Quadratic vs linear
anova(full.model, all.linear) # Full model still better

# 5. Interaction reduction
anova(full.model, full.model.noint)

# So the best model is the one without the traits!
tab_model(full.model)

# Predictions and vizualizations!
traitnm <- "lsla.d"
predictAndVisualize <- function (deltaData, traitnm, n=100) {
  other_traits <- names(deltaData)[!(names(deltaData) %in% c(traitnm, 
                                                             "site", 
                                                             "garden", 
                                                             "treatment", 
                                                             "unified_names",
                                                             "pg.cat",
                                                             "comparison"))]

  focal.column = seq(from = min(deltaData[, traitnm], na.rm = T),
                     to = max(deltaData[, traitnm], na.rm = T), 
                     length = n)
  other.vals <- t(apply(deltaData[, other_traits], 2, mean, na.rm = T))
  preddata <- cbind(focal.column,
        do.call("rbind", replicate(n, other.vals, simplify = FALSE)))
  colnames(preddata)[1] <- traitnm
  preddata <- rbind(preddata, preddata, preddata)
  preddata$site <- rep(c("wanang", "numba", "yawan"), each = n)

  # predvals <- AICcmodavg::predictSE.gls(full.sla.linear.noint.nosite, newdata = preddata,
  #                   newdtype = 'response')
  # preddata$predvals <- predvals$fit
  # preddata$predse <- predvals$se.fit

preddata <- data.frame(lsla.d = seq(from = min(deltaData$lsla.d, na.rm = T), 
                                  to = max(deltaData$lsla.d, na.rm = T), 
                                  length = 100),
                       water.cont = mean(deltaData$water.cont, na.rm = T), 
                       perc.herb= mean(deltaData$perc.herb, na.rm = T), 
                       richness= mean(deltaData$richness, na.rm = T), 
                       labu= mean(deltaData$labu, na.rm = T),
                       lrrsla= mean(deltaData$lrrsla, na.rm = T))

predvals <- AICcmodavg::predictSE.gls(full.sla.linear.noint.nosite, newdata = preddata,
                    newdtype = 'response')
preddata$predvals <- predvals$fit
preddata$predse <- predvals$se.fit
